---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rosettaR

<!-- badges: start -->
<!-- badges: end -->

The goal of rosettaR is to provide a series of tools to work with Rosetta Statements, which are plain language statements that can be converted to semantic data when paired with a template. 

## Installation

You can install the development version of rosettaR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("timalamenciak/rosettaR")
```

## Example

This is a basic example of how `rosettaR` can convert data:

```{r example}
library(rosettaR)
statement <- "Kitchener is located in Canada"
in_template <- "{{ city }} is located in {{ country }}"
out_template <- "CITY,COUNTRY,,{{ city }},{{ country }}"
df <- rosettaFormat(statement, in_template, out_template)
```

# Development notes

This is a very early version of this package. It is based on the idea of Rosetta Statements described in [Rosetta Statements: Simplifying FAIR Knowledge Graph Construction with a User-Centered Approach](https://arxiv.org/abs/2407.20007) by Lars Vogt et al.
